Software variability in fMRI analysis:\\ comparing between-tool and numerical errors

### 2021-12-03 paper/paper.tex
Edits in results

In~\cite{salari2021accurate}, Fuzzy Libmath was shown to accurately
simulate the effect of Linux operating system updates in structural
pre-processing pipelines. To validate our pipeline instrumentations for the present study, we first verified that non-instrumented
executions of the same pipeline on the same dataset led to identical
results \tristan{did you fix random seeds and if so in which pipelines?}. We also listed the pipeline
library dependencies using the \texttt{ldd} Linux utility and verified that
(1) the tested pipelines were dynamically linked to the GNU libmath library, and 
(2) there was no alternative implementation of elementary mathematical functions in the pipeline dependencies.
Finally, we verified that the use of Fuzzy libmath affected computational results.

@@

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
identical checksums. For SPM, the checksums were different but differences
were visually unnoticeable. For AFNI, the activation maps were similar
overall, however, differences were more noticeable visually. 
The observed
differences remained small and might be due to the use of Octave vs MATLAB in
SPM. We performed visual quality control of the AFNI and SPM results
for each individual subject \tristan{and confirmed that ...}.
\tristan{add a figure to show the differences for SPM (Octave) and AFNI}

\subsection{Summary group-level statistics}

Summary statistics for the group-level unthresholded and thresholded
t-statistics are reported in Table~\ref{table:pipeline-stats}. BT
variability for tool pair (A, B) was larger than WT variability for tool A and tool B for any tool pair, in both thresholded and
unthresholded maps (Wilcoxon signed-rank test, t-test
p~\textless~$10^{-163}$ for all tests).
% This was confirmed by computing the T-test and Wilcoxon signed-rank statistics between BT and WT as:

@@

\subsection{Group-level unthresholded maps}
% \subsection{Variability of unthresholded statistical maps}

Figure~\ref{fig:unthresh-maps} compares BT and WT variability in
unthresholded maps. WT variability remains lower than BT variability
overall. Interestingly, BT and WT variability appear correlated in a subset
of voxels scattered across the brain
(Figure~\ref{fig:unthresh-correlation}), but the reason for this apparent
correlation remains unknown.

%%%%%%%%%% Var. of Unthresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-unthresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Unthresholded group-level variability computed between tools (\textbf{A}) and within tools (\textbf{B}), and difference between them (\textbf{C}).}
    \label{fig:unthresh-maps}
    \end{minipage}}
  \end{figure*}

  %%%%%%%%%% Corr. plot of tstats%%%%%%%%
  \begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std-corr-unthresh-plot.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \hfill
      \begin{subfigure}[ht]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/std/correlated-unthresh.png}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
      \end{subfigure}
      \caption{
        Voxel-wise comparison of BT and WT variabilty. The bottom row shows 
        voxels where the BT/WT ratio is in (0.5, 2).}
    \label{fig:unthresh-correlation}
    \end{minipage}}
  \end{figure*}

@@

  Figure~\ref{fig:unthresh-maps-sbj} shows BT and WT variability for
  the subject with the highest WT variability among the 16 subjects. For
  this particular subject, WT variability approaches and even surpasses BT
  variability in some regions. This subject also revealed that WT
  variability mainly concentrated in some regions.

@@

\begin{figure*}[ht]
  \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
    \begin{subfigure}[ht]{\textwidth}
      \centering
      \includegraphics[width=.6\textwidth]{figures/sbj-std-corr-unthresh-plot.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \end{subfigure}
    \caption{
      Voxel-wise comparison of BT and WT variabilty subject by subject.}
  \label{fig:unthresh-correlation-allsbj}
  \end{minipage}}
\end{figure*}

@@

While the previous results were obtained at the virtual precision of 53~bits for double-precision values
and 24~bits for single-precision values, we also evaluated WT variability across different virtual precisions for FSL.
Figure~\ref{fig:across-precisions} shows the root-mean-square error (RMSE) between BT and WT variability maps.
The virtual precision of t=17 bits minimized the RMSE between BT and WT.
Figure~\ref{fig:gnp-mni} shows the corresponding BT and WT maps obtained at t=17 bits, showing 
that BT and WT reach comparable magnitudes in some regions.

@@

  %%%%%%%%%% plot different precisions%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
        \centering
        \includegraphics[width=.75\textwidth]{figures/bg_global_precision.pdf}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
        \caption{Unthresholded group t-statistics standard deviations computed between tools (\textbf{A}) and within tools (\textbf{B}), and difference between them (\textbf{C}).
        WT is computed at the virtual precision of t=17 bits.}
    \label{fig:gnp-mni}
    \end{minipage}}
  \end{figure}

@@

  Figure~\ref{fig:thresh-maps} compares BT and WT for thresholded group-level maps.
  While BT is consistently larger than WT, thresholding is an unstable operation 
  that results in comparable magnitudes for BT and WT within activation clusters where BT variability is low.

  Figure~\ref{fig:cm-thresh} compares regional activation instability for BT and WT. The average ratio
  of unstable regions was 26\% for BT and 9.2\% for WT. Moreover, the average F1 score was 0.28. Therefore, regional 
  WT instability does not appear to be a 
  good predictor of BT instability.

@@

>>>>

In~\cite{salari2021accurate}, Fuzzy Libmath was shown to accurately
simulate the effect of Linux operating system updates in structural
pre-processing pipelines. To validate our pipeline instrumentations for the present study, we first verified that non-instrumented
executions of the same pipeline on the same dataset led to identical
results. We also listed the pipeline
library dependencies using the \texttt{ldd} Linux utility and verified that
(1) the tested pipelines were dynamically linked to the GNU libmath library, and 
(2) there was no alternative implementation of elementary mathematical functions in the pipeline dependencies.
Finally, we verified that the use of Fuzzy libmath affected computational results.

@@

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
identical checksums. For SPM, the checksums were different but differences
were visually unnoticeable. For AFNI, the activation maps were similar
overall, however, differences were more noticeable visually. 
The observed differences remained small and might be due to the use of
Octave vs MATLAB in SPM. We performed visual quality control of the AFNI
and SPM results for each individual subject and confirmed that images were
correctly registered to the MNI template.

\subsection{Group-level unthresholded maps}

Summary statistics for the group-level unthresholded and thresholded
t-statistics are reported in Table~\ref{table:pipeline-stats}. For each
tool pair (A, B), BT variability was significantly larger than machine
error in tool A or B, in both thresholded and unthresholded maps (Wilcoxon
signed-rank test p=0 for all tests, t-test p~\textless~$10^{-163}$ for all
tests).

% This was confirmed by computing the T-test and Wilcoxon signed-rank statistics between BT and WT as:

@@

% \subsection{Variability of unthresholded statistical maps}

Figure~\ref{fig:unthresh-maps} compares BT and WT variability in
unthresholded maps. WT variability remains lower than BT variability
overall. Moreover, BT and WT variability appeared moderately correlated (R
in [0.56, 0.58], p=0.0). The set of voxels exhibiting comparable BT and
WT values (located around the identity line in the XY plot) does not seem to have 
any spatial consistency.

%%%%%%%%%% Var. of Unthresh %%%%%%%%
\begin{figure*}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
      \centering
      \includegraphics[width=\textwidth]{figures/std/gl-unthresh.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \caption{Unthresholded group-level variability computed between tools (\textbf{A}) and within tools (\textbf{B}),
    and difference between them (\textbf{C}). \tristan{Add XY plots for each map}}
    \label{fig:unthresh-maps}
    \end{minipage}}
  \end{figure*}

@@

  Figure~\ref{fig:unthresh-maps-sbj} shows BT and WT variability for
  the subject with the highest WT variability among the 16 subjects. For
  this particular subject, WT variability approaches and even surpasses BT
  variability in some regions. Besides, numerical variability appears focal.

  As for the group maps, BT and WT variability appeared moderately
  correlated for all subjects (Figure~\ref{fig:unthresh-correlation-allsbj}, 
  R in [0.442, 0.612], p=0.0 for all
  subjects).
  \tristan{Add the subject XY plots as supplementary material}

@@

@@

While the previous results were obtained for machine error, we also evaluated numerical variability across different virtual precisions for FSL.
Figure~\ref{fig:across-precisions} shows the root-mean-square error (RMSE) between BT and WT variability maps.
The virtual precision of t=17 bits minimized the RMSE between BT and WT.
Figure~\ref{fig:gnp-mni} shows the corresponding BT and WT maps obtained at t=17 bits, showing 
that BT and WT reach comparable magnitudes in some regions.

@@

  %%%%%%%%%% plot different precisions%%%%%%%%
  \begin{figure}[ht]
    \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
        \centering
        \includegraphics[width=.75\textwidth]{figures/bg_global_precision.pdf}
        %\caption{Standard deviation of thresholded t-statistics map on template surface}
        \caption{Unthresholded group t-statistics standard deviations computed between tools (\textbf{A}) and within tools (\textbf{B}), and difference between them (\textbf{C}).
        WT is computed at the virtual precision of t=17 bits.\tristan{merge with figure with brain maps}}
    \label{fig:gnp-mni}
    \end{minipage}}
  \end{figure}

@@

  Figure~\ref{fig:thresh-maps} compares BT and WT for thresholded group-level maps.
  While BT is consistently larger than machine error, thresholding is an unstable operation 
  that results in comparable magnitudes for BT and WT within activation clusters where BT variability is low.

  Figure~\ref{fig:cm-thresh} compares regional activation instability for BT and WT. The average ratio
  of unstable regions was 26\% for BT and 9.2\% for WT. The average F1 score in these confusion matrics was 0.28. Regional 
  WT instability appears to moderately explain BT regional BT instability.

@@

\appendix

\tristan{Add SPM and AFNI replication figure here.}

\begin{figure*}[ht]
  \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
    \begin{subfigure}[ht]{\textwidth}
      \centering
      \includegraphics[width=.6\textwidth]{figures/sbj-std-corr-unthresh-plot.png}
      %\caption{Standard deviation of thresholded t-statistics map on template surface}
    \end{subfigure}
    \caption{
      Voxel-wise comparison of BT and WT variabilty subject by subject. \tristan{update figure}}
  \label{fig:unthresh-correlation-allsbj}
  \end{minipage}}
\end{figure*}
